---
title: "Partitioning workflow using Allen Coral Atlas data"
format:
  html:
    toc: true
vignette: >
  %\VignetteIndexEntry{Partitioning workflow using Allen Coral Atlas data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

This vignette sets out and example reef partitioning workflow using data from
the [Allen Coral Atlas](https://allencoralatlas.org/). The data intended for use 
are accessible from the Allen Coral Atlas globally.

While this example vignette refers to the Allen Coral Atlas, this approach is 
flexible and can be used on any raster and vector data beyond habitat, bathymetry
and reef examples.

# Required packages

```{r package loading}
#| eval: false

library(ReefPartitionUniversal)
library(tidyverse)
library(terra)
library(sf)
library(ggplot2)
library(RColorBrewer)
library(exactextractr)
```

# Loading and formatting required data

The Allen Coral Atlas is aims to map coral reefs globally and provides accessible
coral reef ecology and environmental data layers accessible [online](https://allencoralatlas.org/).
Using the Allen Coral Atlas online map a reef area can be selected using the `Polygon tool`.
Once selected, data are available for download. The minimum data required for this
workflow are `Reef Extent` (reef outline polygons), `Geomorphic Map` (reef habitat data)
and a continuous data layer (`Bathymetry - composite depth` or `Annual Turbidity`).
In this example we use bathymetry data as the continuous `additional_variable` for
partitioning.

Allen Coral Atlas `Geomorphic Map` and `Benthic Map` data are downloaded in a
vector format (e.g. .gpkg file) containing area polygons. To use these data as
habitat layers for habitat layers to extract points from during analyses, these
layers must be converted to raster data. This can be done using the 
`ReefPartitionUniversal::vector_to_raster` function. As we have existing data for
the workflow (bathymetry data) it is recommended to use that data as a template
for vector -> raster conversion. (If no template is available, 
`ReefPartitionUniversal::generate_raster_template` can be used).



```{r data preparation}
#| eval: false
starting_crs <- 4326
output_crs <- 3857

# Load reef outline vector data (reef extent)
reef_polygons <- st_read(
  "Path to reef extent .gpkg"
)

# Ensure reef polygons are not segmented to ensure full clustering
reef_polygons <- reef_polygons %>%
  st_transform(output_crs) %>%
  summarise(geometry = st_union(geom)) %>%
  st_cast("POLYGON") %>%
  mutate(id = 1:n())

# Load bathymetry raster data from ACA dataset
bathymetry_raster <- rast(
  "Path to reef bathymetry .tif"
) %>%
  project(., str_glue("epsg:{output_crs}"))

# Load geomorphic habitat vector data from ACA dataset
habitat_vector <- st_read(
  "Path to geomorphic habitat .gpkg"
) %>%
  rename(habitat = class) %>%
  st_transform(crs = output_crs)

# Convert habitat data to raster format
habitat_raster <- vector_to_raster(
  habitat_vector,
  "habitat",
  output_file = NA,
  raster_template = bathymetry_raster # Use existing bathymetry raster as the template (i.e. grid format) for raster conversion
)
habitat_levels <- habitat_raster$levels # Store factor levels contained in new `habitat_raster
habitat_raster <- habitat_raster$raster
```

# Define arguments for workflow

The following code defines arguments for use in our example workflow with One
Tree Island Reef. This includes defining the target habitat types from `habitat_raster`,
defining the desired H3 cell resolution for pixel extraction, and the desired
site size for clustering.

```{r defining arguments}
#| eval: false

set.seed(123)

# Select habitat levels that correspond to desired habitat categories
habitat_categories <- habitat_levels[
  habitat_levels$character %in%
    c("Reef Crest", "Reef Slope", "Outer Reef Flat", "Sheltered Reef Slope"),
]$numeric_levels

# Define the desired site size in terms of spatial area
# and the number of H3 cells per site
site_size <- 250 * 250 # set size size for our example to 625,000m~2~
pixel_area <- res(habitat_raster)[1] * res(habitat_raster)[1] # Calculate raster pixel square area (m~2~)
n_pixels <- site_size / pixel_area # Calculate desired number of points per site

# Select a target reef from the reef polygons available
target_reef <- reef_polygons[1, ]
```

# Extracting pixel data for the target reef

The remainder of the workflow using Allen Coral Atlas data follows other reef partitioning
workflows.
Once the input data and arguments have been set up, we can extract pixel data for
our reef using the `target_reef` outline polygon. 
This process involves identifying selected pixels from the habitat raster object
and extracting bathymetry data.

```{r pixel extraction}
#| eval: false
pixel_data <- extract_point_pixels(
  reef_polygon = target_reef,
  habitat_raster = habitat_raster,
  add_var_raster = bathymetry_raster,
  habitat_categories = habitat_categories,
  output_epsg = output_crs
)

# We must also attach a reef ID to the dataframe and
# remove any pixels with invalid depth data
pixel_data$UNIQUE_ID <- "Reef 1"
head(pixel_data)
```

# Clustering habitat pixels based on distance and depth
Data for each required pixel have been extracted, and we can now cluster pixels
based on their geographic distance and depth. Using the default arguments for
`cluster_reef_pixels()` clusters pixels within each habitat type using a Minimum
Spanning Tree and `reef_skater_fast()` skater clustering algorithm implemented 
using igraph. The returned dataframe contains a row for each pixel and an additional
column containing the clustered `site_id`.

```{r cluster pixels}
#| eval: false
clustered_pixels <- cluster_reef_points(
  pixel_data,
  clustering_function_args = list(
    point_area = pixel_area,
    site_size = site_size
  )
)
```

# Creating site polygons
Once pixels for the reef have been assigned site IDs we can collate them into site
polygons and use post-processing to separate site areas that contain large distances
into smaller site IDs.

```{r cluster polygons}
#| eval: false

# Collate H3 cells that are assigned site IDs into polygons
clustered_polygons <- pixels_to_polygons(
  clustered_pixels,
  UNIQUE_ID,
  habitat,
  depth,
  pixel_size = res(habitat_raster)[1]
)

# Site postprocessing with a minimum number of pixels per site of 50
processed_sites <- site_postprocessing(
  clustered_polygons,
  min_site_area = 50 * pixel_area
)
```

# Mapping outputs using ggplot2
```{r mapping}
#| eval: false
# Reorder site ID labels to improve readability
sampled_ids <- sample(levels(processed_sites$site_id))
processed_sites$sampled_id <- factor(
  processed_sites$site_id,
  levels = sampled_ids
)

# Plot site polygons coloured by IDs (colours are repeated)
id_map <- ggplot() +
  geom_sf(data = processed_sites, aes(fill = sampled_id)) +
  theme(legend.position = "none") # Remove legend due to too many site ID labels

# Re-extract depth data for site polygons for plotting
processed_sites$depth <- abs(exact_extract(
  bathymetry_raster,
  processed_sites,
  "mean"
))

# Plot sites coloured by mean depth
depth_map <- ggplot() +
  geom_sf(data = processed_sites, aes(fill = depth)) +
  scale_fill_fermenter(
    palette = "greens",
    direction = 1
  )
```